---
title: "Miscellaneous functions"
author: "Manuel Gimond"
format: 
  html:
    code-overflow: wrap
    toc: true
editor: visual
editor_options: 
  chunk_output_type: console
---

------------------------------------------------------------------------

```{r echo = FALSE, message=FALSE, warning=FALSE}
options(message = FALSE, warning = FALSE)
library(knitr)
opts_knit$set(global.par = TRUE)
```

```{r echo = FALSE}
par(mar = c(0, 0, 0, 0))
```

# Geospatial functions

### Generate a uniform grid of points

```{r}
source("https://raw.githubusercontent.com/mgimond/miscR/master/geo_grid_unif.R")
```

The `geo_grid_unif()` function will generate a grid of points on a sphere. By default, the output is a matrix of longitude and latitude values. If an `sf` object is desired, the output will be a point `sf` object in a WGS84 coordinate system.

Arguments:

-   `n`: Number of points
-   `sf`: Logical indicating if an `sf` object should be created (default output is a matrix)

Example:

```{r fig.width = 4, fig.height = 4, message=FALSE, warning=FALSE}
#| code-fold: true
library(sf)

source("https://raw.githubusercontent.com/mgimond/miscR/master/geo_grid_unif.R")

# Generate uniform points
p <- geo_grid_unif(n =100, sf = TRUE)

# Download world layer for reference
world <-  readRDS(gzcon(url("https://github.com/mgimond/tissot/raw/master/smpl_world.rds")))

# Reproject to "world from space" (this CRS makes it easier to view
# the uniform distribution of points on a sphere)
world_sp <- st_transform(world, crs = "ESRI:102038")
p_sp <- st_transform(p, crs = "ESRI:102038")

## Plot the data
plot(p_sp, pch = 16, col = "darkred", graticule = st_graticule(world_sp ))
plot(st_geometry(world_sp), add = TRUE, border = "grey")
```

------------------------------------------------------------------------

### Generate point buffers on an azimuthal equidistant CS

```{r}
source("https://raw.githubusercontent.com/mgimond/miscR/master/geo_buf.R")
```

The `geo_buf()` function generates geometrically correct buffers from an `sf` point object stored in a geographic coordinate system. Output consists of an `sf` polygon object in a WGS84 coordinate system. The buffers are generated in ESRI's *World Azimuthal Equidistant* projection centered on each point whose buffer is generated.

Arguments:

-   `p`: `sf` point object in a geographic coordinate system (i.e. lon/lat values).
-   `seg`: Number of line segments for a quarter of the circle (e.g. a seg = 3 will generate a circle with 12 line segments). The default is `30`.
-   `radius`: Buffer radius in meters. Default is `5000` meters.

Example:

```{r fig.width = 4, fig.height = 4, message=FALSE, warning=FALSE}
#| code-fold: true
library(sf)

source("https://raw.githubusercontent.com/mgimond/miscR/master/geo_buf.R")

# Download world layer for reference
world <-  readRDS(gzcon(url("https://github.com/mgimond/tissot/raw/master/smpl_world.rds")))

# Create point layer
p <- data.frame( y = c(rep( seq(-60,60, by = 30), times = 5), 90),
                 x = c(rep( seq(-150, 0, by = 35), each = 5 ), 0))
p.sf <- st_as_sf(p, coords = c("x","y"), crs = "EPSG:4326")

# Create buffers
buf <- geo_buf(p.sf, radius = 500000)

# Reproject to "world from space" (this CRS makes it easier to view
# the uniform distribution of points on a sphere)
world_sp <- st_transform(world, crs = "ESRI:102038")
buf_sp <- st_transform(buf, crs = "ESRI:102038")

## Plot the data
plot(buf_sp, pch = 16, col = rgb(1,0,0,0.3), 
     graticule = st_graticule(world_sp ))
plot(st_geometry(world_sp), add = TRUE, border = "grey")
```

Note that the function does not generate a true geodesic buffer. It generates one from an equidistant projection (*World Azimuthal Equidistant*) centered on each point in the input dataset. As such, error increases with increasing buffer radius.

```{r echo = FALSE}
par(mar = c(4, 4, 0, 1))
```

```{r, fig.height=2.7, fig.width=3.9}
#| code-fold: true
buf_area <- function(pt, x) {
  (geo_buf(pt, radius = x) |> st_area() ) / (x^2 * pi) * 100
}
rad <- seq(1,9500000, length.out = 10)
pt <- st_point(c(-70,45)) |> st_sfc(crs = "EPSG:4326") |> st_as_sf()
err <- 100 - sapply(rad, FUN = \(x) buf_area(pt, x) )

plot(rad / 1000, abs(err), xlab = "Radius (km)", ylab = "% error (log scale)", 
     col = "red", pch = 16, cex = 1.5, log = "y",  las =1, yaxt="n")
lab <- seq(1,15, by= 2)
axis(2, at = lab,  label = lab, las = 1)
abline(h=lab, lty = 3, col = "grey")
```

```{r echo = FALSE}
par(mar = c(0, 0, 0, 0))
```

### Generate geodesic point buffers on a sphere

```{r}
source("https://raw.githubusercontent.com/mgimond/miscR/master/geo_buf_geodesic.R")
```

The `geo_buf_geodesic()` function generates geodesic buffers from an `sf` point object stored in a geographic coordinate system. Output consists of an `sf` polygon object in a WGS84 coordinate system. The buffers are generated on a sphere.

Arguments:

-   `p`: `sf` point object in a geographic coordinate system (i.e. lon/lat values).
-   `seg`: Number of line segments for a quarter of the circle (e.g. a seg = 3 will generate a circle with 12 line segments). The default is `30`.
-   `radius`: Buffer radius in meters. Default is `5000` meters.
-   `sphere.r` : Sphere radius in meters (default is  `6371007.2` m).

Example:

```{r fig.width = 4, fig.height = 4, message=FALSE, warning=FALSE}
#| code-fold: true
library(sf)

source("https://raw.githubusercontent.com/mgimond/miscR/master/geo_buf_geodesic.R")

# Download world layer for reference
world <-  readRDS(gzcon(url("https://github.com/mgimond/tissot/raw/master/smpl_world.rds")))

# Create point layer
p <- data.frame( y = c(rep( seq(-60,60, by = 30), times = 5), 90),
                 x = c(rep( seq(-150, 0, by = 35), each = 5 ), 0))
p.sf <- st_as_sf(p, coords = c("x","y"), crs = "EPSG:4326")

# Create buffers
buf <- geo_buf_geodesic(p.sf, radius = 500000)

# Reproject to "world from space" (this CRS makes it easier to view
# the uniform distribution of points on a sphere)
world_sp <- st_transform(world, crs = "ESRI:102038")
buf_sp <- st_transform(buf, crs = "ESRI:102038")

## Plot the data
plot(buf_sp, pch = 16, col = rgb(1,0,0,0.3), 
     graticule = st_graticule(world_sp ))
plot(st_geometry(world_sp), add = TRUE, border = "grey")
```

The `geo_buf_geodesic()` does not improve on areal error measurements over its `geo_buf()` counterpart at large radial distances, but it does benefit from computational improvements over `geo_buf()` with a processing time about 20 times faster.
